geochemistry

### 2025-03-18 geochemistry.qmd
Macbook work session

## DB setup
pxrf_raw <- pxrf_raw |> 
  filter(!Sample_ID %in% c("I19-1751", "H21-2126")) # Removing 2 altered samples
pxrf_analysis <- pxrf_raw |> 
  select("Sample_ID", "pXRF_ID", "Formation/Type",
         "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "Fe2O3", "Cu",
         "Sr", "Zr") |>  # Removed elements with 0 or <LOD values in all samples or in high percentages in all groups
  mutate(`Formation/Type` = case_when(
    `Formation/Type` == "TYPE 1" ~ "T1",
    `Formation/Type` == "TYPE 2" | `Formation/Type` == "TYPE 2B" | `Formation/Type` == "TYPE 2R" ~ "T2",
    `Formation/Type` == "TYPE 3" ~ "T3",
    `Formation/Type` == "TYPE 4" ~ "T4",
    `Formation/Type` == "TYPE 5" ~ "T5",
    `Formation/Type` == "TYPE 6" | `Formation/Type` == "TYPE 6LLB" | `Formation/Type` == "TYPE 6E" ~ "T6",
    `Formation/Type` == "TYPE 7" ~ "T7",
    `Formation/Type` == "TYPE 8" ~ "T8",
    `Formation/Type` == "TYPE 9" ~ "T9",
    `Formation/Type` == "TYPE 10" ~ "T10",
    `Formation/Type` == "TYPE 11" ~ "T11",
    TRUE ~ `Formation/Type`
  ))

@@

# Table to see the occurrence of 0 or <LOD values within each element and by rm type, to choose the elements in which to apply statistics.
# Combine the multiple columns into a single table
lod_tab <- pxrf_raw |>
  select(-contains("Err")) |>
  select(-pXRF_ID, -Application, -Method, -ElapsedTime, -`Cal Check`) |>
  mutate(across(.cols = MgO:U,
                .fns = ~ case_when(
                  . == 0 ~"Success 0",
                  . > 0 ~ "Success",
                  TRUE ~ as.character(.))
                ))
# Reshape the data into a long format
lod_tab_long <- lod_tab |>
  pivot_longer(cols = MgO:U,  # Assuming these columns contain the 'Success' and 'LOD' values
               names_to = "Element",  # New column for the names of elements
               values_to = "Value")   # New column for the corresponding Success/LOD values

@@

# Now, reshape the data into a wide format using pivot_wider
lod_tab_table <- lod_tab_summary |>
  pivot_wider(names_from = Value, values_from = Count, values_fill = list(Count = 0))  # Fill missing values with 0

lod_tab_summary <- lod_tab_long |>
  group_by(Element, Value) |>
  summarise(Count = n(), .groups = "drop")  # Count occurrences of each Value per Element

lod_tab_summary <- lod_tab_summary |>
  count(Element, Value)

# Create the bar plot
ggplot(lod_tab_summary, aes(x = Element, y = Count, fill = Value)) +
  geom_bar(stat = "identity", position = "stack") +  # Use stat = "identity" to use summarized counts
  scale_fill_manual(values = c("Success" = "darkseagreen", "Success 0" = "darkseagreen2", "< LOD" = "indianred3")) +
  labs(title = "Summarized Success and LOD Counts by Element",
       y = "Count", 
       fill = " ") +
  theme_minimal() +
 # facet_wrap(~ Element, scales = "free_y") +  # Facet by Element with independent y-axes
  theme(axis.text.x = element_text(hjust = 1),
        axis.title.x = element_blank()) +
  #geom_text(data = lod_tab_summary_filtered, aes(label = Count), position=position_stack(vjust=0.5), colour="white", size = 3) +
  coord_flip() +
  scale_y_continuous(breaks = seq(0, max(lod_tab_summary$Count), by = 20))

@@

pxrf_average <- pxrf_analysis |> 
  select(-Sample_ID, -pXRF_ID) |>
  mutate(`Formation/Type` = case_when(
    `Formation/Type` == "T2" | `Formation/Type` == "T3" | `Formation/Type` == "T4" | `Formation/Type` == "T5" ~ "T2-5", TRUE ~ `Formation/Type`
  )) |> 
  mutate(`Formation/Type` = factor(`Formation/Type`, levels = c("TW", "LJW", "MJE", "UJW", "UJE", "CPT", "UCA", "T1", "T2-5", "T6", "T7", "T8", "T9", "T10", "T11"))) |>
  mutate(across(-`Formation/Type`, ~ as.numeric(.))) |>    # Convert valid numbers
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), 0, .))) |>  # Replace NAs with 0
  group_by(`Formation/Type`) |> 
  summarise(across(where(is.numeric),
                   .fns = list(mean = ~ round(mean(.x, na.rm = TRUE), 4), 
                               sd = ~ round(sd(.x, na.rm = TRUE), 4)), 
                   .names = "{.col}_{.fn}"))

@@

# Boxplot data preparation
boxplots_pxrf <- pxrf_analysis |> 
  select(-Sample_ID, -pXRF_ID) |> 
  filter(!`Formation/Type` %in% c("TW", "T1")) |> # Removing chalcedony
  mutate(across(-`Formation/Type`, ~ as.numeric(.)))
  #pivot_longer(cols = c("MgO", "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "Fe2O3", "Cu",
        # "Sr", "Zr"), names_to = "Variable", values_to = "Value")
boxplots_pxrf <- boxplots_pxrf |> 
  mutate(`Formation/Type` = case_when(
    `Formation/Type` == "T2" | `Formation/Type` == "T3" | `Formation/Type` == "T4" | `Formation/Type` == "T5" ~ "T2-5", TRUE ~ `Formation/Type`
  )) |> 
  mutate(`Formation/Type` = factor(`Formation/Type`, levels = c("TW", "LJW", "MJE", "UJW", "UJE", "CPT", "UCA", "T1", "T2-5", "T6", "T7", "T8", "T9", "T10", "T11")))

@@

b3 <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = SiO2, colour = `Formation/Type`)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  geom_text(data = count_data_sio2, aes(x = `Formation/Type`, y = 70, label = n), size = 3) +
  coord_cartesian(ylim = c(70, 100)) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

b8 <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = CaO, colour = `Formation/Type`)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  geom_text(data = count_data_cao, aes(x = `Formation/Type`, y = -0.3, label = n), size = 3) +
  theme_minimal()

@@

b9 <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = TiO2, colour = `Formation/Type`)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  geom_text(data = count_data_tio2, aes(x = `Formation/Type`, y = -0.02, label = n), size = 3) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

b10 <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = Fe2O3, colour = `Formation/Type`)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  geom_text(data = count_data_fe2o3, aes(x = `Formation/Type`, y = 0, label = n), size = 3) +
  coord_cartesian(ylim = c(0, 0.75)) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

boxplots_comb1 <- ggarrange(b3, b10, b9, b2, b4, b7, 
          labels = c("a","b","c","d","e","f"),
          ncol = 2,
          nrow = 3,
          common.legend = TRUE,
          legend = "bottom")
boxplots_comb2 <- ggarrange(b5, b11, b12, b13, 
          labels = c("a","b","c","d"),
          ncol = 2,
          nrow = 2,
          common.legend = TRUE,
          legend = "bottom")

>>>>

## DB setup
pxrf_db <- pxrf_raw |> 
  filter(!Sample_ID %in% c("I19-1751", "H21-2126")) # Removing 2 altered samples
pxrf_analysis <- pxrf_db |> 
  select("Sample_ID", "pXRF_ID", "Formation/Type",
         "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "Fe2O3", "Cu",
         "Sr", "Zr") |>  # Removed elements with 0 or <LOD values in all samples or in high percentages in all groups
  mutate(`Formation/Type` = case_when(
    `Formation/Type` == "TYPE 1" ~ "T1",
    `Formation/Type` == "TYPE 2" | `Formation/Type` == "TYPE 2B" | `Formation/Type` == "TYPE 2R" ~ "T2",
    `Formation/Type` == "TYPE 3" ~ "T3",
    `Formation/Type` == "TYPE 4" ~ "T4",
    `Formation/Type` == "TYPE 5" ~ "T5",
    `Formation/Type` == "TYPE 6" | `Formation/Type` == "TYPE 6LLB" | `Formation/Type` == "TYPE 6E" ~ "T6",
    `Formation/Type` == "TYPE 7" ~ "T7",
    `Formation/Type` == "TYPE 8" ~ "T8",
    `Formation/Type` == "TYPE 9" ~ "T9",
    `Formation/Type` == "TYPE 10" ~ "T10",
    `Formation/Type` == "TYPE 11" ~ "T11",
    TRUE ~ `Formation/Type`
  ))

@@

# Table to see the occurrence of 0 or <LOD values within each element and by rm type, to choose the elements in which to apply statistics.
# Combine the multiple columns into a single table
lod_tab <- pxrf_db |>
  select(-contains("Err")) |>
  select(-pXRF_ID, -Application, -Method, -ElapsedTime, -`Cal Check`, -Sample_ID) |>
  mutate(across(.cols = MgO:U,
                .fns = ~ case_when(
                  . == 0 ~"Success 0",
                  . > 0 ~ "Success",
                  TRUE ~ as.character(.))
                ))
# Reshape the data into a long format
lod_tab_long <- lod_tab |>
  pivot_longer(cols = MgO:U,  # Assuming these columns contain the 'Success' and 'LOD' values
               names_to = "Element",  # New column for the names of elements
               values_to = "Value")   # New column for the corresponding Success/LOD values

@@

# Reshape the data into a wide format using pivot_wider
lod_tab_table <- lod_tab_summary |>
  pivot_wider(names_from = Value, values_from = Count, values_fill = list(Count = 0)) |> 
  filter(Success > 80) |> 
  select(Element, Success, "< LOD")
lod_tab_table_long <- lod_tab_table |> 
  pivot_longer(cols = c(Success, `< LOD`), names_to = "Metric", values_to = "Count") |> 
  mutate(Metric = factor(Metric, levels = c("Success", "< LOD"))) 
lod_tab_table_long |> 
  ggplot(aes(x = Element, y = Count, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of Success and < LOD", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("Success" = "#99D3A7", "< LOD" = "#ED7673"))

# Create the bar plot
ggplot(lod_tab_summary, aes(x = Element, y = Count, fill = Value)) +
  geom_bar(stat = "identity", position = "stack") +  # Use stat = "identity" to use summarized counts
  scale_fill_manual(values = c("Success" = "darkseagreen", "Success 0" = "lightgoldenrod1", "< LOD" = "indianred3")) +
  labs(title = "Summarized Success and LOD Counts by Element",
       y = "Count", 
       fill = " ") +
  theme_minimal() +
 # facet_wrap(~ Element, scales = "free_y") +  # Facet by Element with independent y-axes
  theme(axis.text.x = element_text(hjust = 1),
        axis.title.x = element_blank()) +
  #geom_text(data = lod_tab_summary_filtered, aes(label = Count), position=position_stack(vjust=0.5), colour="white", size = 3) +
  coord_flip() +
  scale_y_continuous(breaks = seq(0, max(lod_tab_summary$Count), by = 20))

@@

pxrf_average <- pxrf_analysis |> 
  select(-Sample_ID, -pXRF_ID) |>
  mutate(`Formation/Type` = case_when(
    `Formation/Type` == "T2" | `Formation/Type` == "T3" | `Formation/Type` == "T4" | `Formation/Type` == "T5" ~ "T2-5", TRUE ~ `Formation/Type`
  )) |> 
  mutate(`Formation/Type` = factor(`Formation/Type`, levels = c("TW", "LJW", "MJE", "UJW", "UJE", "CPT", "UCA", "T1", "T2-5", "T6", "T7", "T8", "T9", "T10", "T11"))) |>
  mutate(across(-`Formation/Type`, ~ as.numeric(.))) |>    # Convert valid numbers
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), 0, .))) |>  # Replace NAs with 0
  group_by(`Formation/Type`) |> 
  summarise(across(where(is.numeric),
                   .fns = list(mean = ~ round(mean(.x, na.rm = TRUE), 4), 
                               sd = ~ round(sd(.x, na.rm = TRUE), 4),
                               count = ~ sum(.x != 0, na.rm = TRUE)),
                   .names = "{.col}_{.fn}"))

@@

# Boxplot data preparation
boxplots_pxrf <- pxrf_analysis |> 
  select(-Sample_ID, -pXRF_ID) |> 
  filter(!`Formation/Type` %in% c("TW", "T1", "T9", "T10", "T11")) |> # Removing chalcedony
  mutate(across(-`Formation/Type`, ~ as.numeric(.)))
  #pivot_longer(cols = c("MgO", "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "Fe2O3", "Cu",
        # "Sr", "Zr"), names_to = "Variable", values_to = "Value")
boxplots_pxrf <- boxplots_pxrf |> 
  mutate(`Formation/Type` = case_when(
    `Formation/Type` == "T2" | `Formation/Type` == "T3" | `Formation/Type` == "T4" | `Formation/Type` == "T5" ~ "T2-5", TRUE ~ `Formation/Type`
  )) |> 
  mutate(`Formation/Type` = factor(`Formation/Type`, levels = c("TW", "LJW", "MJE", "UJW", "UJE", "CPT", "UCA", "T1", "T2-5", "T6", "T7", "T8", "T9", "T10", "T11")))

@@

b3 <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = SiO2, colour = `Formation/Type`)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_sio2, aes(x = `Formation/Type`, y = 70, label = n), size = 3) +
  coord_cartesian(ylim = c(70, 100)) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

b8 <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = CaO, colour = `Formation/Type`)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_cao, aes(x = `Formation/Type`, y = -0.3, label = n), size = 3) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

b9 <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = TiO2, colour = `Formation/Type`)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_tio2, aes(x = `Formation/Type`, y = -0.02, label = n), size = 3) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

b10 <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = Fe2O3, colour = `Formation/Type`)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_fe2o3, aes(x = `Formation/Type`, y = 0, label = n), size = 3) +
  coord_cartesian(ylim = c(0, 0.75)) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

boxplots <- ggarrange(b3, b8, b10, b9, 
          labels = c("a","b","c","d"),
          ncol = 2,
          nrow = 2,
          common.legend = TRUE,
          legend = "bottom")
boxplots_comb2 <- ggarrange(b5, b11, b12, b13, 
          labels = c("a","b","c","d"),
          ncol = 2,
          nrow = 2,
          common.legend = TRUE,
          legend = "bottom")
